---
title: "Eigenvalues and Eigenvectors: The Hidden Engines of Linear Dynamics"
subtitle: "Why every economist should make friends with the eigendecomposition"
author: "Scarlett Xie"
date: "Sep 13,2026"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    code-link: true
execute:
  warning: false
  message: false
  fig-width: 7
  fig-height: 5
  fig-dpi: 150
---

```{r}
#| label: setup
#| include: false

# Install ggplot2 only if it is not already available -------------------------
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2", repos = "https://cloud.r-project.org")
}
library(ggplot2)

# Global plotting style and output location ------------------------------------
theme_set(theme_bw(base_size = 12))
knitr::opts_chunk$set(fig.path = "figures/", echo = FALSE)
set.seed(20260913)
```

# Background {#sec-question}

Most relationships that economists put on a computer are, at their core, **linear**: regressions, vector autoregressions (VARs), Markov transition matrices, state-space models, and even the local approximations used to solve dynamic stochastic general equilibrium (DSGE) models. Every one of these tools repeatedly multiplies vectors by matrices.

A natural question follows:

> *When we multiply by the same matrix over and over, what happens in the long run — and why?*

The complete answer is given by two objects that live inside the matrix itself: its **eigenvalues** and **eigenvectors**. This post explains what they are, how to compute them by hand and in R, and why they do so much of the heavy lifting in economics — from the long-run unemployment rate to whether an estimated macro model is even stable.

# What are eigenvalues and eigenvectors? {#sec-definition}

Multiplying a vector $\mathbf{x}$ by a matrix $\mathbf{A}$ generally does two things at once: **rotates** the vector and **stretches** it. An **eigenvector** is a special, non-zero direction that is *not* rotated at all — it is only stretched (or shrunk, or flipped). The amount of stretching along that direction is the corresponding **eigenvalue**.

The defining equation is short:

$$
\mathbf{A}\mathbf{v} = \lambda \mathbf{v}, \qquad \mathbf{v} \neq \mathbf{0},
$$

where

-   $\mathbf{A}$ is an $n \times n$ square matrix,
-   $\mathbf{v}$ is an eigenvector of $\mathbf{A}$,
-   $\lambda$ (lambda) is the associated eigenvalue, which may be positive, negative, zero, or even complex.

In words: *along the direction* $\mathbf{v}$, the action of the entire matrix reduces to multiplication by a single number $\lambda$.

## The characteristic equation

Moving $\lambda\mathbf{v}$ to the left and inserting the identity matrix $\mathbf{I}$ gives

$$
(\mathbf{A} - \lambda\mathbf{I})\mathbf{v} = \mathbf{0}.
$$

A non-zero solution $\mathbf{v}$ exists exactly when the matrix $(\mathbf{A}-\lambda\mathbf{I})$ is singular, i.e. its determinant is zero:

$$
\det(\mathbf{A} - \lambda\mathbf{I}) = 0.
$$

This is the **characteristic equation**. It produces a polynomial in $\lambda$ whose $n$ roots are the eigenvalues; substituting each root back into $(\mathbf{A}-\lambda\mathbf{I})\mathbf{v} = \mathbf{0}$ gives the eigenvectors.

## A 2 × 2 example

Take the symmetric matrix

$$
\mathbf{A} = \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}.
$$

The characteristic equation is

$$
\det\begin{pmatrix} 2-\lambda & 1 \\ 1 & 2-\lambda \end{pmatrix}
= (2-\lambda)^2 - 1
= \lambda^2 - 4\lambda + 3
= (\lambda - 1)(\lambda - 3) = 0,
$$

so the eigenvalues are $\lambda_1 = 3$ and $\lambda_2 = 1$. The eigenvectors (defined only up to scale, so conventionally normalized to unit length) are

$$
\lambda_1 = 3: \quad \mathbf{v}_1 = \frac{1}{\sqrt 2}\begin{pmatrix}1\\1\end{pmatrix},
\qquad
\lambda_2 = 1: \quad \mathbf{v}_2 = \frac{1}{\sqrt 2}\begin{pmatrix}1\\-1\end{pmatrix}.
$$

Quick check: along the $45^\circ$ direction, $\mathbf{A}(1,1)' = (3,3)' = 3(1,1)'$, so the matrix stretches that direction threefold; along $(1,-1)'$ it leaves the vector exactly unchanged.

# Graphical representation {#sec-picture}

The definition becomes much easier to *see*. The code below draws the unit circle, applies $\mathbf{A}$ to every point on it (the circle becomes an ellipse), and overlays the two eigenvectors. Every point on the circle is both rotated and stretched **except** points on the eigenvector directions, which stay on their own lines.

```{r}
#| label: fig-geometric
#| fig-cap: "The unit circle and its image under A.  Most directions are rotated and stretched, but the eigenvector directions (red, green) are only stretched, by factors of 3 and 1."

A <- matrix(c(2, 1, 1, 2), nrow = 2, byrow = TRUE)

# Points on the unit circle, and their image under A ---------------------------
theta <- seq(0, 2 * pi, length.out = 400)
unit_circle <- rbind(cos(theta), sin(theta))
image <- A %*% unit_circle

shapes <- rbind(
  data.frame(x = unit_circle[1, ], y = unit_circle[2, ],
             set = "Unit circle"),
  data.frame(x = image[1, ], y = image[2, ],
             set = "Image under A (ellipse)")
)

# Eigenvectors as unit vectors; Av = lambda v ---------------------------------
v_lambda3 <- c(1, 1) / sqrt(2)   # stretched to length 3
v_lambda1 <- c(1, -1) / sqrt(2)  # unchanged (Av = v), so only one arrow shown
Av3 <- as.numeric(A %*% v_lambda3)  # A%*%v is 2x1; extract both coordinates

segments <- data.frame(
  x = 0, y = 0,
  xend = c(v_lambda3[1], Av3[1], v_lambda1[1]),
  yend = c(v_lambda3[2], Av3[2], v_lambda1[2]),
  direction = c("v,  lambda = 3", "Av = 3v,  lambda = 3",
                "v = Av,  lambda = 1")
)

ggplot(shapes, aes(x = x, y = y, colour = set)) +
  geom_path(linewidth = 0.9) +
  geom_segment(
    data = segments,
    aes(x = x, y = y, xend = xend, yend = yend, colour = direction),
    arrow = arrow(length = grid::unit(0.14, "inches")),
    linewidth = 0.9, inherit.aes = FALSE
  ) +
  geom_hline(yintercept = 0, linewidth = 0.2) +
  geom_vline(xintercept = 0, linewidth = 0.2) +
  coord_fixed() +
  scale_x_continuous(limits = c(-3.4, 3.4)) +
  scale_y_continuous(limits = c(-3.4, 3.4)) +
  scale_colour_manual(
    name = NULL,
    values = c("grey60", "#08519c", "#de2d26", "#a50f15", "#31a354")
  ) +
  labs(x = expression(x[1]), y = expression(x[2]),
       title = "Eigenvectors are the directions that do not rotate",
       subtitle = "Unit circle (grey) mapped to an ellipse (blue) by A")
```

The ellipse's principal axes line up exactly with $\mathbf{v}_1$ and $\mathbf{v}_2$, and its semi-axis lengths are $3$ and $1$ — precisely the eigenvalues. An eigen-decomposition is, geometrically, a description of the shape the matrix draws.

# Why eigenvalues govern *dynamics* {#sec-dynamics}

The really important payoff is not one multiplication but **repeated** multiplication. Consider the linear dynamical system

$$
\mathbf{x}_{t+1} = \mathbf{A}\mathbf{x}_t
\quad\Longrightarrow\quad
\mathbf{x}_t = \mathbf{A}^t \mathbf{x}_0.
$$

Computing $\mathbf{A}^t$ directly is painful. Instead, collect the eigenvectors as columns of $\mathbf{P}$ and the eigenvalues in a diagonal matrix $\mathbf{D}$:

$$
\mathbf{A} = \mathbf{P}\mathbf{D}\mathbf{P}^{-1},
\qquad
\mathbf{D} = \begin{pmatrix}\lambda_1 & 0\\0 & \lambda_2\end{pmatrix}.
$$

This is the **diagonalization**. Powers now simplify dramatically, because all the cross-terms in $\mathbf{P}^{-1}\mathbf{P}$ cancel:

$$
\mathbf{A}^t = \mathbf{P}\mathbf{D}^t\mathbf{P}^{-1},
\qquad
\mathbf{D}^t = \begin{pmatrix}\lambda_1^t & 0\\0 & \lambda_2^t\end{pmatrix}.
$$

Expanding the initial condition in the eigenvector basis, $\mathbf{P}^{-1}\mathbf{x}_0 = \mathbf{c}$, gives

$$
\mathbf{x}_t = \mathbf{P}\mathbf{D}^t\mathbf{c}
= \sum_{i=1}^{n} c_i \lambda_i^t \mathbf{v}_i.
$$

This single formula explains the long run:

-   any component with $|\lambda_i| < 1$ **decays geometrically** toward zero;
-   a component with $|\lambda_i| > 1$ **explodes**;
-   a component with $\lambda_i = 1$ **persists forever** (a steady state, or — when unwanted — a unit root);
-   a negative or complex $\lambda_i$ produces oscillations (complex pairs rotate with period set by their angle).

The eigenvalues are therefore not an abstract exam question. They *are* the dynamics.

# Economic application 1: a labor-market Markov chain {#sec-markov}

Suppose each month workers move between employment ($E$) and unemployment ($U$) according to the transition matrix

$$
\mathbf{P} = \begin{pmatrix} 0.90 & 0.10 \\ 0.25 & 0.75 \end{pmatrix},
$$

where rows are *today's* state and columns are *next month's* state: an employed worker stays employed with probability $0.90$, and an unemployed worker finds a job with probability $0.25$.

If $\boldsymbol{\pi}_t$ is the row vector of population shares, then

$$
\boldsymbol{\pi}_{t+1} = \boldsymbol{\pi}_t \mathbf{P}.
$$

The **stationary distribution** $\boldsymbol{\pi}$ satisfies $\boldsymbol{\pi} = \boldsymbol{\pi}\mathbf{P}$ with entries summing to one. Transposing, $\mathbf{P}'\boldsymbol{\pi}' = 1\cdot\boldsymbol{\pi}'$: the stationary distribution is exactly the **left eigenvector of** $\mathbf{P}$ associated with eigenvalue 1, normalized to sum to one. This is why every stationary Markov chain has an eigenvalue of one.

```{r}
#| label: markov-setup

P <- matrix(c(0.90, 0.10,
              0.25, 0.75),
            nrow = 2, byrow = TRUE,
            dimnames = list(c("Employed", "Unemployed"),
                            c("Employed", "Unemployed")))
P
```

Solving on paper first: stationarity requires $0.10\,\pi_E = 0.25\,\pi_U$, so $\pi_E = 2.5\,\pi_U$, and $\pi_E + \pi_U = 1$ gives

$$
\pi_E = \frac{5}{7} \approx 0.714, \qquad \pi_U = \frac{2}{7} \approx 0.286.
$$

Now let R do the same job through the eigen decomposition (left eigenvectors of $\mathbf{P}$ are ordinary eigenvectors of $\mathbf{P}'$):

```{r}
#| label: markov-eigen

eig_P <- eigen(t(P))
eig_P$values

# Eigenvector paired with eigenvalue 1, renormalized to sum to one ------------
stationary <- eig_P$vectors[, 1] / sum(eig_P$vectors[, 1])
round(stationary, 4)

# Verify it really is stationary ----------------------------------------------
round(stationary %*% P, 4)
```

The second eigenvalue, $\lambda_2 = 0.65$, governs how fast a workforce *off* its stationary composition converges back to it: the gap shrinks like $0.65^t$ — roughly a $35\%$ reduction per month. We can see this by iterating the chain from several very different starting distributions.

```{r}
#| label: fig-convergence
#| fig-cap: "Convergence to the stationary distribution from three different initial states. Solid lines = employed share; dotted lines = unemployed share. All paths approach 5/7 and 2/7, the eigenvector for lambda = 1."

horizon <- 20
initial_states <- list(
  c(1, 0),       # everyone starts employed
  c(0.2, 0.8),   # deep recession: 80% unemployed
  c(0, 1)        # everyone starts unemployed
)

paths <- lapply(seq_along(initial_states), function(i) {
  dist <- matrix(NA_real_, nrow = horizon + 1, ncol = 2)
  dist[1, ] <- initial_states[[i]]
  for (t in seq_len(horizon)) {
    dist[t + 1, ] <- dist[t, ] %*% P
  }
  data.frame(
    t = 0:horizon,
    employed = dist[, 1],
    unemployed = dist[, 2],
    start = paste0("Initial employed share = ", initial_states[[i]][1])
  )
})
convergence <- do.call(rbind, paths)

pi_E <- 5 / 7
pi_U <- 2 / 7

ggplot(convergence, aes(x = t)) +
  geom_hline(yintercept = c(pi_E, pi_U), linetype = "dashed",
             colour = "grey50") +
  geom_line(aes(y = employed, colour = start), linewidth = 0.9) +
  geom_line(aes(y = unemployed, colour = start), linewidth = 0.9,
            linetype = "dotted") +
  annotate("text", x = horizon, y = pi_E, label = " stationary E: 71.4%",
           hjust = 1, vjust = -0.6, size = 3.4, colour = "grey30") +
  annotate("text", x = horizon, y = pi_U, label = " stationary U: 28.6%",
           hjust = 1, vjust = 1.6, size = 3.4, colour = "grey30") +
  scale_x_continuous(breaks = seq(0, horizon, 5)) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "Months elapsed", y = "Share of the labor force",
       colour = NULL,
       title = "Where does the labor market settle?",
       subtitle = "Ergodicity: the long run forgets where it started") +
  theme(legend.position = "bottom")
```

The economic reading is substantive: regardless of whether everyone starts with a job or $80\%$ start unemployed, the *same* long-run unemployment rate, about $28.6\%$, emerges. Initial conditions are forgotten because the non-unit eigenvalue lies strictly inside the unit circle. This "forgetfulness" property — called **ergodicity** — is what licenses treating a single economy's long-run averages as population-level statements, and it is diagnosed entirely from eigenvalues.

# Economic application 2: stability of dynamic models {#sec-stability}

The same mathematics answers a question every macro forecaster has to ask after estimating a VAR: *is the estimated system stable?*

For a vector autoregression written in companion form, $\mathbf{z}_{t+1} = \mathbf{C}\mathbf{z}_t$, a one-time shock dies out if and only if

$$
|\lambda_i(\mathbf{C})| < 1 \quad \text{for every } i,
$$

i.e. every eigenvalue of the companion matrix lies **inside the unit circle** in the complex plane. An eigenvalue outside the circle means a small perturbation grows without bound; an eigenvalue on the circle means a unit root and permanent effects. (In a continuous-time system $\dot{\mathbf{x}} = \mathbf{A}\mathbf{x}$, the equivalent condition is $\operatorname{Re}(\lambda_i) < 0$.)

The code below plots the spectrum of the labor-market transition matrix against the unit circle — the exact graphical stability check reported in software such as EViews or `vars::stability`.

```{r}
#| label: fig-unit-circle
#| fig-cap: "Eigenvalues of the transition matrix in the complex plane. The eigenvalue lambda = 1 anchors the stationary distribution; lambda = 0.65 sits inside the unit circle, so all deviations from stationarity decay. A stable VAR requires every eigenvalue of its companion matrix to lie strictly inside this circle."

theta <- seq(0, 2 * pi, length.out = 400)
circle <- data.frame(x = cos(theta), y = sin(theta))

spectrum <- data.frame(
  x = Re(eig_P$values),
  y = Im(eig_P$values)
)

# 标签位置：lambda = 1 放上方，lambda = 0.65 放下方，避免重叠
label_pos <- data.frame(
  x = c(1.00, 0.65),
  y = c(0.20, -0.22),
  label = c("lambda == 1", "lambda == 0.65")
)

ggplot() +
  geom_polygon(data = circle, aes(x = x, y = y),
               fill = "#6baed6", alpha = 0.18) +
  geom_path(data = circle, aes(x = x, y = y), linetype = "dashed") +
  geom_hline(yintercept = 0, linewidth = 0.2) +
  geom_vline(xintercept = 0, linewidth = 0.2) +
  geom_point(data = spectrum, aes(x = x, y = y),
             size = 3.5, colour = "#cb181d") +
  geom_text(data = label_pos,
            aes(x = x, y = y, label = label),
            parse = TRUE, size = 3.8, colour = "#cb181d") +
  annotate("text", x = -0.25, y = 0, label = "|lambda| < 1:\nshocks decay",
           size = 3.6, colour = "#08519c") +
  coord_fixed() +
  scale_x_continuous(limits = c(-1.35, 1.35)) +
  scale_y_continuous(limits = c(-1.35, 1.35)) +
  labs(x = "Re(lambda)", y = "Im(lambda)",
       title = "The unit-circle stability test")
```

Here $\lambda = 1$ sits *on* the circle — exactly as it must, since the level of employment persists — while deviations from that level, governed by $0.65$, decay. In a VAR, by contrast, a unit eigenvalue signals a non-stationary process that must be differenced or cointegrated before standard inference is valid.

# More exploration {#sec-other-uses}

The eigen decomposition recurs throughout empirical and theoretical work:

-   **Principal components analysis (PCA).** PCs are the eigenvectors of the covariance (or correlation) matrix, ordered by eigenvalue; the eigenvalue is the variance captured by that component. Macro "factors" extracted from hundreds of time series, and the widely used FRED-MD data set, are built this way.
-   **PageRank and network centrality.** Google's original PageRank is the stationary distribution — again the eigenvector for eigenvalue one — of a Markov chain over web pages. The same construction measures input-output influence and systemic risk in financial networks.
-   **Local dynamics of nonlinear models.** Linearizing a DSGE model around its steady state produces a matrix whose eigenvalues determine saddle-path stability: some eigenvalues must be explosive (forward-looking "jump" variables) and others stable for a unique equilibrium to exist.
-   **Quadratic forms and GLS.** Weighted distance, efficient GMM, and the geometry of multivariate regression all simplify by rotating data onto eigenvector axes.

# Programming basis:R {#sec-r-summary}

Everything above used only base R's `eigen()`, which returns a list with `$values` and a `$vectors` matrix whose *columns* are the eigenvectors:

Two practical notes. First, eigenvectors are only identified up to sign and scale — `eigen()` may return $-\mathbf{v}$ where a textbook prints $\mathbf{v}$; the span is identical. Second, real matrices can have complex conjugate eigenvalue pairs (when $\mathbf{A}$ is not symmetric), and `eigen()` will store them as complex numbers automatically; plot $\operatorname{Re}(\lambda)$ against $\operatorname{Im}(\lambda)$ exactly as in @fig-unit-circle.

# Takeaways {#sec-takeaways}

1.  **Eigenvectors are the directions a matrix leaves unrotated; eigenvalues are the stretch factors along them**, $\mathbf{A}\mathbf{v} = \lambda\mathbf{v}$.
2.  **Diagonalization turns repeated multiplication into scalar powers**: $\mathbf{A}^t\mathbf{x}_0 = \sum_i c_i\lambda_i^t\mathbf{v}_i$, so eigenvalues completely determine long-run behavior — decay, persistence, explosion, or oscillation.
3.  **Economics runs on this machinery**: a stationary Markov distribution is the eigenvector for $\lambda = 1$; its other eigenvalues set convergence speed; and VAR/DSGE stability is a unit-circle (or half-plane) test on eigenvalues.
4.  **In practice one call to `eigen()` does it** — the conceptual work is knowing *which* eigenvalue answers the question and *how to read it*.

Whenever a model can be written as repeated multiplication by a matrix, the question "what happens in the long run?" always has the same address: its eigenvalues.

# Further reading {#sec-reading}

-   Strang, G. *Introduction to Linear Algebra* (chapters on eigenvalues and Markov matrices).
-   Simon, C. P., & Blume, L. *Mathematics for Economists* (eigenvalues, stability, and phase diagrams).
-   Hamilton, J. D. *Time Series Analysis* (VAR companion-form stability and unit roots).
