---
title: "When Young Children Are at Home, Who Is at Work?"
subtitle: "The employment divide between women and men in the 2025 CPS"
author: "Scarlett Xie"
date: 2026-09-28
categories: [Labor economics, IPUMS CPS, Data visualization]
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    fig-dpi: 180
execute:
  warning: false
  message: false
  echo: true
---

<style>
body {font-family: Georgia, "Times New Roman", serif; color:#24343b}
h1,h2,h3,.quarto-title-block {font-family: Arial, Helvetica, sans-serif}
.quarto-title-block .title {color:#152d3e; max-width:850px}
#quarto-content {max-width:1100px; margin:auto}
main.content {max-width:780px; line-height:1.62; font-size:1.03rem}
figcaption {font: .87rem Arial, Helvetica, sans-serif; color:#52616b}
</style>

```{r}
#| label: setup-and-download
#| results: hide
# Run this document in RStudio/Quarto. An IPUMS CPS account and API key are
# needed on the first render. Cached data can be used without the key later.
# Step 1: check R packages. The actual API key lives in your private R session.
required <- c("ipumsr", "dplyr", "srvyr", "survey", "ggplot2", "readr", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  message("Installing missing R packages: ", paste(missing, collapse = ", "))
  tryCatch(
    utils::install.packages(missing, repos = "https://cloud.r-project.org",
                            type = if (.Platform$OS.type == "windows") "binary" else "source"),
    error = function(e) message("Automatic package installation failed: ", conditionMessage(e))
  )
  missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) {
    stop("These packages are still missing: ", paste(missing, collapse = ", "),
         ". In the RStudio Console, run install.packages(c(",
         paste(sprintf('"%s"', missing), collapse = ", "),
         "), repos = 'https://cloud.r-project.org'), then render again.", call. = FALSE)
  }
}
# Quarto evaluates code in the document directory by default. In some knitr
# versions current_input() names an intermediate file rather than this QMD.
blog_dir <- normalizePath(getwd(), winslash = "/")
# Step 2: create paths. Keep licensed microdata outside the website/repository.
raw_dir <- file.path(path.expand("~"), ".ipums-cps-blog3-cache")
derived_dir <- file.path(blog_dir, "data", "derived")
figure_dir <- file.path(blog_dir, "files")
for (d in c(raw_dir, derived_dir, figure_dir))
  dir.create(d, recursive = TRUE, showWarnings = FALSE)

# Step 3: use a complete local extract when available, otherwise request it.
ddi_files <- list.files(raw_dir, pattern = "^cps_[0-9]+[.]xml$", full.names = TRUE)
complete_ddi <- ddi_files[
  file.exists(sub("[.]xml$", ".csv.gz", ddi_files)) |
    file.exists(sub("[.]xml$", ".dat.gz", ddi_files))
]
if (length(complete_ddi) > 1L)
  stop("More than one complete CPS extract is in the private cache; keep only the 2025 ASEC extract.")
if (length(complete_ddi) == 1L) {
  ddi_path <- complete_ddi[[1]]
} else {
  if (!nzchar(Sys.getenv("IPUMS_API_KEY"))) {
    stop("Set IPUMS_API_KEY in your own RStudio session and register for IPUMS CPS. ",
         "See the replication section.", call. = FALSE)
  }
  request_path <- file.path(raw_dir, "extract_request.rds")
  ddi_path <- tryCatch({
    if (file.exists(request_path)) {
      submitted <- ipumsr::get_extract_info(readRDS(request_path))
    } else {
      extract <- ipumsr::define_extract_micro(
        collection = "cps",
        description = "Blog 3: young children and employment, 2025 ASEC",
        samples = "cps2025_03s",
        variables = c("YEAR", "AGE", "SEX", "YNGCH", "EMPSTAT", "EDUC",
                      "ASECWT", "REPWTP"),
        data_format = "csv", data_structure = "rectangular"
      )
      submitted <- ipumsr::submit_extract(extract)
      saveRDS(submitted, request_path)
    }
    ready <- ipumsr::wait_for_extract(submitted)
    ipumsr::download_extract(ready, download_dir = raw_dir, overwrite = TRUE)
  }, error = function(e) {
    if (grepl("Invalid IPUMS registration|not registered to IPUMS cps",
              conditionMessage(e), ignore.case = TRUE)) {
      stop("IPUMS API says the account associated with this R session's API key ",
           "does not have an active CPS registration. Check that the email in the ",
           "API error matches the email of your CPS website login and the account ",
           "where you copied the key: https://account.ipums.org/api_keys . ",
           "If they match and CPS says you are already registered, contact ",
           "IPUMS User Support at ipums@umn.edu with the 401 error (never send the key). ",
           "Original API message: ", conditionMessage(e), call. = FALSE)
    }
    stop(e)
  })
}
# Step 4: read the DDI and its accompanying compressed CSV into R.
cps_raw <- ipumsr::read_ipums_micro(ddi_path)
rep_names <- grep("^REPWTP[0-9]+$", names(cps_raw), value = TRUE)
stopifnot(all(c("YEAR", "AGE", "SEX", "YNGCH", "EMPSTAT", "EDUC", "ASECWT") %in%
              names(cps_raw)), length(rep_names) == 160L,
          all(as.numeric(cps_raw$YEAR) == 2025L))
```

When a young child needs care, someone has to provide it or pay for it. That
constraint matters for household budgets, employers and the design of child
care policy. Yet a simple claim about "parents' employment" can hide two
quite different patterns. This post asks: **how does employment differ between
adults with a young child at home and adults without a minor child at home,
and is the comparison the same for women and men?**

The 2025 US Current Population Survey offers a useful snapshot. It cannot
follow the same parents before and after a birth, but it can show where the
labor market differences lie. I start with employment, then separate labor
force participation from unemployment, and finally check whether broad age
and education differences account for the contrast.

## The employment split

```{r}
#| label: clean-and-survey-design
#| results: hide
# Step 5: recode child age, employment and controls; retain eligible adults.
# EMPSTAT general code 1 = employed; 2 = unemployed; 3 = not in labor force.
# YNGCH 99 = no identified own child living in the household.
# EDUC 999 = missing or unknown. EDUC 000 may also mean no schooling, so retain it.
cps <- cps_raw |>
  dplyr::mutate(
    sex = factor(as.numeric(SEX), levels = c(1, 2), labels = c("Men", "Women")),
    child_group = dplyr::case_when(
      YNGCH == 99 | YNGCH >= 18 ~ "no_minor",
      YNGCH >= 0 & YNGCH <= 5 ~ "age0_5",
      YNGCH >= 6 & YNGCH <= 12 ~ "age6_12",
      YNGCH >= 13 & YNGCH <= 17 ~ "age13_17",
      TRUE ~ NA_character_
    ),
    child_group = factor(child_group,
                         levels = c("no_minor", "age0_5", "age6_12", "age13_17")),
    status = floor(as.numeric(EMPSTAT) / 10),
    employed = as.numeric(status == 1),
    labor_force = as.numeric(status %in% c(1, 2)),
    unemployed = as.numeric(status == 2),
    age_band = factor(dplyr::case_when(
      AGE <= 34 ~ "25-34", AGE <= 44 ~ "35-44", TRUE ~ "45-54"
    ), levels = c("25-34", "35-44", "45-54")),
    college = factor(ifelse(EDUC >= 111, "BA or higher", "Below BA"),
                     levels = c("Below BA", "BA or higher"))
  ) |>
  dplyr::filter(AGE >= 25, AGE <= 54, !is.na(sex), !is.na(child_group),
                status %in% c(1, 2, 3), !is.na(EDUC), EDUC != 999,
                !is.na(ASECWT), ASECWT > 0)
if (!nrow(cps)) stop("No eligible records: inspect IPUMS variable codes.")
# Audit the sample restriction. These are unweighted person records.
n_downloaded <- nrow(cps_raw)
n_prime_age <- sum(!is.na(cps_raw$AGE) & cps_raw$AGE >= 25 & cps_raw$AGE <= 54)
n_analytic <- nrow(cps)
stopifnot(n_analytic <= n_prime_age, n_prime_age <= n_downloaded)
# Step 6: make the ASEC person-weighted survey design with all 160 replicates.
design <- srvyr::as_survey_rep(
  cps, weight = ASECWT, repweights = dplyr::matches("^REPWTP[0-9]+$"),
  type = "successive-difference", mse = TRUE
)
# Step 7: estimate eight group rates and their design-based confidence intervals.
rates <- design |>
  dplyr::group_by(sex, child_group) |>
  dplyr::summarise(
    employment = srvyr::survey_mean(employed, vartype = "ci", na.rm = TRUE),
    participation = srvyr::survey_mean(labor_force, vartype = "ci", na.rm = TRUE),
    unemployment = srvyr::survey_ratio(unemployed, labor_force,
                                        vartype = "ci", na.rm = TRUE),
    .groups = "drop"
  ) |>
  as.data.frame()
readr::write_csv(rates, file.path(derived_dir, "weighted_rates.csv"))
pick_rate <- function(s, g, col = "employment") {
  row <- rates[rates$sex == s & rates$child_group == g, , drop = FALSE]
  if (nrow(row) != 1L) stop("An expected group is missing: ", s, ", ", g)
  as.numeric(row[[col]])
}
e_w_young <- pick_rate("Women", "age0_5")
e_m_young <- pick_rate("Men", "age0_5")
e_w_none <- pick_rate("Women", "no_minor")
e_m_none <- pick_rate("Men", "no_minor")
women_gap <- 100 * (e_w_young - e_w_none)
men_gap <- 100 * (e_m_young - e_m_none)
fmt_pct <- function(x) sprintf("%.1f%%", 100 * x)
fmt_pp <- function(x) sprintf("%+.1f percentage points", x)
# Consistency check: employment = participation * (1 - unemployment).
stopifnot(max(abs(rates$employment - rates$participation *
                    (1 - rates$unemployment))) < 1e-8)
```

Among adults aged 25 to 54, the employment rate for women whose youngest own
child at home is aged 0 to 5 is **`r fmt_pct(e_w_young)`**. For women with no
own minor child at home, it is `r fmt_pct(e_w_none)`. Among men, the
corresponding rates are **`r fmt_pct(e_m_young)`** and `r fmt_pct(e_m_none)`.
The comparison runs in opposite directions: `r fmt_pp(women_gap)` for women
and `r fmt_pp(men_gap)` for men. Each bar is a CPS person-weighted population
estimate, rather than a count of surveyed people.


```{r}
#| label: fig-employment
#| fig-cap: "Figure 1. Employment rate among adults aged 25-54 by recorded sex and age of youngest own child at home. 2025 IPUMS CPS ASEC; person-weighted with ASECWT. Confidence intervals are in the reproducible rates file."
#| fig-width: 8.5
#| fig-height: 5.6
#| out-width: "100%"
chart_colors <- c(Men = "#315d72", Women = "#ad583f")
child_labels <- c(no_minor = "No minor child at home", age0_5 = "Youngest: 0-5",
                  age6_12 = "Youngest: 6-12", age13_17 = "Youngest: 13-17")
chart_theme <- function() {
  ggplot2::theme_minimal(base_size = 11, base_family = "sans") +
    ggplot2::theme(panel.grid.major.y = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_line(color = "#e5e9eb", linewidth = .3),
      strip.text = ggplot2::element_text(face = "bold", size = 11),
      axis.text = ggplot2::element_text(color = "#334650"),
      plot.margin = ggplot2::margin(10, 14, 8, 8))
}
fig1_data <- dplyr::mutate(rates,
  family = factor(child_labels[as.character(child_group)],
                  levels = rev(unname(child_labels))),
  value = 100 * employment, label = sprintf("%.1f%%", value))
p1 <- ggplot2::ggplot(fig1_data,
  ggplot2::aes(x = family, y = value, fill = sex)) +
  ggplot2::geom_col(width = .66, show.legend = FALSE) +
  ggplot2::geom_text(ggplot2::aes(y = value - 1.7, label = label),
                     hjust = 1, color = "white", fontface = "bold", size = 3.3) +
  ggplot2::facet_wrap(~ sex, ncol = 2) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = chart_colors) +
  ggplot2::scale_y_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100),
                              labels = function(x) paste0(x, "%"), expand = c(0, 0)) +
  ggplot2::labs(y = "Employed, share of all adults", x = NULL) + chart_theme()
ggplot2::ggsave(file.path(figure_dir, "figure1_employment.png"), p1,
                width = 8.5, height = 5.6, dpi = 220, bg = "white")
p1
```

These bars compare different households in one survey year. They do not trace
any family's employment as a child gets older. Nor does the reference group
mean "never had children": it includes people whose children are grown or
live elsewhere.

## The difference shows up in labor force participation

```{r}
#| label: labor-force-comparisons
#| results: hide
lf_w_young <- pick_rate("Women", "age0_5", "participation")
lf_w_none <- pick_rate("Women", "no_minor", "participation")
u_w_young <- pick_rate("Women", "age0_5", "unemployment")
u_w_none <- pick_rate("Women", "no_minor", "unemployment")
```

The employment rate can fall because fewer adults look for paid work, because
more job seekers cannot find work, or both. Figure 2 places everyone in one of
three mutually exclusive states. Among women with a child aged 0 to 5 at home,
`r fmt_pct(lf_w_young)` are in the labor force, versus
`r fmt_pct(lf_w_none)` in the reference group. Their unemployment rates,
measured **among labor force participants**, are `r fmt_pct(u_w_young)` and
`r fmt_pct(u_w_none)`. The conspicuous employment difference therefore sits
mainly alongside a participation difference, not a higher unemployment rate.

```{r}
#| label: fig-labor-force
#| fig-cap: "Figure 2. Employment, unemployment and nonparticipation as shares of all adults in each group. The unemployed segment is a share of all adults, not the unemployment rate (whose denominator is the labor force). 2025 IPUMS CPS ASEC, person-weighted."
#| fig-width: 8.5
#| fig-height: 5.7
#| out-width: "100%"
states <- dplyr::bind_rows(
  dplyr::transmute(rates, sex, child_group, state = "Employed", share = employment),
  dplyr::transmute(rates, sex, child_group, state = "Unemployed", share = participation - employment),
  dplyr::transmute(rates, sex, child_group, state = "Outside labor force", share = 1 - participation)
) |>
  dplyr::mutate(state = factor(state, levels = c("Employed", "Unemployed", "Outside labor force")),
    family = factor(child_labels[as.character(child_group)], levels = rev(unname(child_labels))))
stopifnot(all(states$share >= -1e-8))
p2 <- ggplot2::ggplot(states, ggplot2::aes(x = family, y = 100 * share, fill = state)) +
  ggplot2::geom_col(width = .66) +
  ggplot2::facet_wrap(~ sex, ncol = 2) +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(values = c(Employed = "#315d72", Unemployed = "#d4a25f",
                                        `Outside labor force` = "#d8dee1"), name = NULL) +
  ggplot2::scale_y_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100),
                              labels = function(x) paste0(x, "%"), expand = c(0, 0)) +
  ggplot2::labs(y = "Share of all adults", x = NULL) + chart_theme() +
  ggplot2::theme(legend.position = "bottom", legend.text = ggplot2::element_text(size = 9))
ggplot2::ggsave(file.path(figure_dir, "figure2_labor_force.png"), p2,
                width = 8.5, height = 5.7, dpi = 220, bg = "white")
p2
```

## Does the divide remain after adjustment?

The groups differ in more than family circumstances. For a descriptive check,
I estimate a survey-weighted linear probability model. The outcome is one if a
person is employed and zero otherwise. Let $F_i$ indicate a woman, $C_{ig}$
indicate the youngest-child group $g$ (relative to no minor child at home),
$A_i$ denote age-band indicators and $B_i$ indicate a bachelor's degree or
higher. The estimated equation is

$$
\Pr(E_i=1 \mid F_i,C_i,A_i,B_i)
= \alpha + \beta F_i + \sum_g \gamma_g C_{ig}
+ \sum_g \delta_g(F_i C_{ig}) + \boldsymbol{\theta}' A_i + \lambda B_i.
$$

For the 0-to-5 group, $\gamma_{0-5}$ is the adjusted difference for men;
$\gamma_{0-5}+\delta_{0-5}$ is the difference for women; and
$\delta_{0-5}$ is the difference between those two differences. The reference
person is a man aged 25 to 34, without a bachelor's degree and without an
own minor child at home. Model coefficients are **percentage-point** changes
when multiplied by 100, not percentage changes.

```{r}
#| label: model-and-contrasts
#| results: hide
model <- survey::svyglm(
  employed ~ sex * child_group + age_band + college,
  design = design, family = stats::gaussian()
)
b <- stats::coef(model)
V <- stats::vcov(model)
ci_critical <- stats::qt(.975, df = 159) # 160 person replicates, 159 design df.
men_term <- "child_groupage0_5"
interaction_term <- "sexWomen:child_groupage0_5"
stopifnot(all(c(men_term, interaction_term) %in% names(b)))
contrast <- function(label, terms) {
  cvec <- stats::setNames(rep(0, length(b)), names(b))
  cvec[terms] <- 1
  estimate <- sum(cvec * b)
  se <- sqrt(as.numeric(t(cvec) %*% V %*% cvec))
  data.frame(group = label, estimate_pp = 100 * estimate,
             lower_pp = 100 * (estimate - ci_critical * se),
             upper_pp = 100 * (estimate + ci_critical * se), se_pp = 100 * se)
}
adjusted <- dplyr::bind_rows(
  contrast("Men", men_term),
  contrast("Women", c(men_term, interaction_term)),
  contrast("Women minus men", interaction_term)
)
readr::write_csv(adjusted, file.path(derived_dir, "adjusted_gaps.csv"))
capture.output(summary(model), file = file.path(derived_dir, "model_summary.txt"))
capture.output(sessionInfo(), file = file.path(derived_dir, "session_info.txt"))
adj_w <- adjusted$estimate_pp[adjusted$group == "Women"]
adj_m <- adjusted$estimate_pp[adjusted$group == "Men"]
interaction <- adjusted[adjusted$group == "Women minus men", ]
```

```{r}
#| label: model-table
#| echo: true
# The same model contrasts drive the table, Figure 3 and the prose below.
model_table <- dplyr::mutate(adjusted,
  Comparison = c("Men: 0-5 vs no minor", "Women: 0-5 vs no minor",
                 "Women minus men: difference in differences"),
  `Estimate (pp)` = sprintf("%+.1f", estimate_pp),
  `95% CI (pp)` = sprintf("[%+.1f, %+.1f]", lower_pp, upper_pp))
knitr::kable(model_table[, c("Comparison", "Estimate (pp)", "95% CI (pp)")],
             caption = "Selected contrasts from the weighted linear probability model")
```

### The adjusted employment divide

Figure 3 plots the first two rows of the table. Each dot is an estimated
difference from adults of the same recorded sex with no minor child at home;
the horizontal line is its 95% confidence interval. Zero means no adjusted
difference. The third table row compares the two dots statistically and is
the young-child interaction coefficient. Showing it in the table avoids
mistaking that interaction for a third employment rate.

```{r}
#| label: fig-adjusted
#| fig-cap: "Figure 3. Model-based differences in employment relative to adults of the same recorded sex with no own minor child at home. Age-band and education adjusted; points and 95% replicate-weight intervals match the first two table rows. Units are percentage points."
#| fig-width: 8.5
#| fig-height: 3.6
#| out-width: "100%"
fig3_data <- dplyr::filter(adjusted, group %in% c("Women", "Men")) |>
  dplyr::mutate(group = factor(group, levels = c("Men", "Women")))
axis_max <- ceiling(max(abs(c(fig3_data$lower_pp, fig3_data$upper_pp))) / 5) * 5 + 2
p3 <- ggplot2::ggplot(fig3_data, ggplot2::aes(x = estimate_pp, y = group, color = group)) +
  ggplot2::geom_vline(xintercept = 0, color = "#788991", linewidth = .6) +
  ggplot2::geom_errorbar(ggplot2::aes(xmin = lower_pp, xmax = upper_pp),
                         orientation = "y", width = .13, linewidth = .85) +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_color_manual(values = chart_colors, guide = "none") +
  ggplot2::scale_x_continuous(limits = c(-axis_max, axis_max),
                              labels = function(x) sprintf("%+g", x)) +
  ggplot2::labs(x = "Adjusted difference (percentage points)", y = NULL) + chart_theme()
ggplot2::ggsave(file.path(figure_dir, "figure3_adjusted_gaps.png"), p3,
                width = 8.5, height = 3.6, dpi = 220, bg = "white")
p3
```

The adjusted difference is `r fmt_pp(adj_w)` for women and
`r fmt_pp(adj_m)` for men. The difference between them is
`r fmt_pp(interaction$estimate_pp)` (95% interval:
`r sprintf("%+.1f", interaction$lower_pp)` to
`r sprintf("%+.1f", interaction$upper_pp)` percentage points). The opposite
signs persist after these controls. That is a sharper description of the
employment divide, not an estimate of what a birth would do to a particular
person's job prospects.

## Takeaway
> **A divide the data can reveal, but not explain**

The survey reveals two distinct labor market patterns around young children:
women in the 0-to-5 household group are less likely to be employed than women
without a minor child at home, while men in the corresponding group are more
likely to be employed. For women, participation rather than unemployment is
the more informative margin. The difference survives broad age and education
adjustment. It points to an important question for child care and work policy:
what combination of care costs, household specialization, earnings and
selection into parenthood lies behind it? This cross-sectional comparison
cannot distinguish those mechanisms or support a causal claim.

## Behind the numbers: Data and replication

> **Data note.** The API downloads the full 2025 CPS Annual Social and Economic
> Supplement person extract, including children and older adults. It contains
> `r format(n_downloaded, big.mark = ",")` **unweighted person records**. After
> restricting to ages 25-54, `r format(n_prime_age, big.mark = ",")` remain;
> after removing invalid or missing analysis fields and nonpositive person
> weights, `r format(n_analytic, big.mark = ",")` remain. The latter is a
> subset of the former, not a count of the US population. Population rates
> use the ASEC person weight.

The youngest-child measure identifies a person's youngest *own* child at home,
including biological, step and adopted children as defined by IPUMS. "No minor
child at home" may include people with adult children or children living
elsewhere; it is not synonymous with never having children. The CPS sex field
records male or female; it does not measure gender identity. These details
matter when interpreting the comparison.

| Source field | Purpose |
|---|---|
| AGE, SEX, YNGCH | Age restriction, recorded sex and youngest own child at home |
| EMPSTAT, EDUC | Employment status and education control |
| ASECWT | ASEC person weight for population rates and the regression |
| REPWTP1 through REPWTP160 | Person replicate weights for survey uncertainty |

The QMD requests sample `cps2025_03s` using the IPUMS API, caches the licensed
person data privately outside the website, recodes the variables, computes
weighted rates, fits the interaction model and saves the following outputs.
The source is [IPUMS CPS](https://cps.ipums.org/cps/), based on the Census
Bureau and Bureau of Labor Statistics survey; see the
[youngest-child definition](https://cps.ipums.org/cps-action/variables/YNGCH),
[person weights](https://cps.ipums.org/cps/sample_weights.shtml) and
[replicate-weight guidance](https://cps.ipums.org/cps/repwt.shtml).

| Output | Contents |
|---|---|
| Private home-directory cache | XML dictionary, extract request and person records; never publish |
| `data/derived/weighted_rates.csv` | Eight weighted group rates and intervals |
| `data/derived/adjusted_gaps.csv` | Three regression contrasts and intervals |
| `data/derived/model_summary.txt`, `session_info.txt` | Full regression output and R environment |
| `files/figure1_employment.png` | Employment rates |
| `files/figure2_labor_force.png` | Labor market composition |
| `files/figure3_adjusted_gaps.png` | Adjusted model contrasts |

To replicate, register for IPUMS CPS, set your private API key in the
IPUMS_API_KEY environment variable, restart RStudio and render this QMD.
The first run requests an extract; later runs use the cached copy. The
[README](README.md) gives the folder structure and commands. The full R code
is in the folded code blocks above. Never publish the key or raw person
records. All displayed figures, table entries and numerical statements are
computed from the current extract during render.
